Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-07 text/manuscript-full-KS.qmd
methods revisions

We assigned each census tract to a College Board Geomarket, which we constructed through manual digitization of maps published by the College Board in 2023 marketing materials promoting their Enrollment Planning Service (EPS) software products. Specifically, we georeferenced the College Board’s Geomarket maps by overlaying PDF images onto spatial coordinates aligned with census tract boundaries. We then extracted vector data by manually tracing the Geomarket boundaries and generating individual shapefiles for each Geomarket.

This approach to constructing Geomarkets aligned precisely with 2020 Decennial Census tract boundaries; that is, each census tract corresponded uniquely to a Geomarket generated from the College Board images. However, this process was less precise when applied to the 1980 and 2000 Decennial Census tract boundaries, likely a factor of census tract delineations shifting over time based on demographic changes and our use of map images from 2023. For instance, census tracts that experience population growth exceeding the 8,000 resident threshold can be split into two or more smaller census tracts to maintain consistent population ranges. Conversely, two or more adjacent census tracts experiencing population declines can sometimes be merged to form a single census tract whose population falls within the 2,500 to 8,000 resident range.

Some census tract boundaries from the 1980 and 2000 Decennial Censuses intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets), resulting in spatial misalignment. For tracts divided across multiple Geomarkets, we assigned each tract to the Geomarket that encompassed the majority of its spatial area. This method introduced slight differences to Geomarket shapes across decennial censuses. However, to assess the robustness of our approach to spatial misalignments, we tested alternative methods for resolving ambiguous tract assignments to Geomarkets and found overall results remained relatively consistent across approaches.[^17]

@@

Next, we utilized the st_intersection() function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. We split census-tract shapefiles that intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets) into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people.
--> 
```

**RQ2**. Our second research question asks, how does the racial and socioeconomic composition of included versus excluded prospects vary when student list purchases filter on Geomarkets? We answer this question using actual student lists purchased by public universities. In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from these search criteria.

@@

>>>>

We assigned each census tract to a College Board Geomarket, which we constructed through manual digitization of maps published by The College Board in 2023 marketing materials promoting their Enrollment Planning Service (EPS) software products. Specifically, we georeferenced the College Board’s Geomarket maps by overlaying PDF images onto spatial coordinates aligned with census tract boundaries. We then extracted vector data by manually tracing the Geomarket boundaries and generating individual shapefiles for each Geomarket. For instance, @fig-geomarket-methods shows The College Board's map of Geomarkets in the Chicago metropolitan area on the left and our created shapefiles on the right.

This approach to constructing Geomarkets aligned precisely with 2020 census tract boundaries from ACS 5-year estimates; that is, each census tract corresponded uniquely to a Geomarket generated from the College Board images. For example, @fig-geomarket-misalign zooms into the northern Geomarkets of the Chicago metropolitan area from @fig-geomarket-methods. The center map showcases how our created Geomarket shapefiles (delineated in purple) align with the spatial coordinates of 2020 census tract boundaries (delineated in grey). However, this process was less precise when applied to the 1980 and 2000 Decennial Census tract boundaries, likely a factor of census tract delineations shifting over time based on demographic changes and our use of map images from 2023. The map on the right in @fig-geomarket-misalign deilneates 1980 census tract boundaries, which are visibly larger and more spatially aggregated than the smaller, more granular 2020 tracts shown in the center map.

<!--# census tracts that experience population growth exceeding the 8,000 resident threshold can be split into two or more smaller census tracts to maintain consistent population ranges. Conversely, two or more adjacent census tracts experiencing population declines can sometimes be merged to form a single census tract whose population falls within the 2,500 to 8,000 resident range. -->

Some census tract boundaries from the 1980 and 2000 Decennial Censuses intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets), resulting in spatial misalignment. In such cases, we assigned each tract to the Geomarket that encompassed the majority of its spatial area. When the spatial majority was relatively balanced across multiple Geomarkets, we assigned tracts in a way that maintained Geomarket boundaries that most smoothly followed Geomarket delineations sourced from The College Board. While this approach minimized ambiguity, it introduced minor variations in Geomarket shapes across decennial censuses. We illustrate this process in @fig-geomarket-misalign, using a 1980 census tract in Chicago as an example, where our constructed shapefiles for the Chain of Lakes and North Shore Geomarkets do not align with the tract boundary (circled in red). In this case, because the majority of the tract’s spatial area fell within the Chain of Lakes Geomarket, we assigned the tract to Chain of Lakes rather than to North Shore. This assignment, along with others like it, slightly alters the overall boundaries of Geomarkets when compared to those based on 2020 census tract delineations. To assess the robustness of our approach to spatial misalignments, we tested alternative methods for resolving ambiguous tract assignments to Geomarkets and found overall patterns in the results remained relatively consistent across approaches.[^17]

@@

Next, we utilized the st_intersection() function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. We split census-tract shapefiles that intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets) into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people.
--> 
```
**RQ2**. Our second research question asks, how does the racial and socioeconomic composition of included versus excluded prospects vary when student list purchases filter on Geomarkets? We answer this question using actual student lists purchased by public universities. In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from these search criteria.

@@

```{r fig-geomarket-methods, echo=FALSE,out.width="49%", fig.cap="caption",fig.align='center',fig.cap = "Chicago, College Board and Recreated Geomarkets"}
knitr::include_graphics('./../assets/images/eps/chicago_geos.png')

```{r fig-geomarket-misalign, echo=FALSE,out.width="100%", fig.cap="caption",fig.align='center',fig.cap = "Chicago, College Board and Recreated Geomarket Spatial Misalignment"}
knitr::include_graphics('./../assets/images/eps/chicago_misalign.png')
```
